Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-27 write-ups/FSE/paper.tex
got the paper to fitmake

We discuss our experience porting irregular algorithms to cooperative
kernels and describe the GPUs on which we evaluate these applications
(\mysec\ref{sec:portingalgorithms}).  For these GPUs, we report on
experiments to determine non-cooperative workloads that model the
requirements of various graphics rendering tasks
(\mysec\ref{sec:sizingnoncoop}).  We then examine the overhead
associated with moving to cooperative kernels when multitasking is
\emph{not} required (\mysec\ref{sec:overhead}), as well as the
responsiveness and throughput observed when a cooperative kernel is
multi-tasked with non-cooperative workloads
(\mysec\ref{sec:responsiveness}).

To our knowledge, there are no competing methods against which to
meaningfully compare our approach.  We are not aware of any existing
platform that supports multitasking between long-running, fully-occupant
OpenCL kernels.  Very recent \nvidia{} GPUs, to which we do not yet have
access, support preemption for graphics
applications~\cite{PascalWhitepaper}, but our understanding is that this
cannot yet be leveraged from CUDA, and certainly not from OpenCL.

@@

We devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgroups of a
GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
searched for a maximum value for $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering.  Using the computed value, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period and the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
period at which the OS requires the GPU for display rendering,
respectively.

@@

Our results are the following: light workloads needs to run 3ms every
70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
the medium and heavy workloads, the 40ms period corresponds to graphics
rendering occurring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders. We use these results to guide our evaluation of the
responsiveness of cooperative scheduling in
\mysec\ref{sec:responsiveness}.

@@

Our experiments span 11 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of
workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 240 configurations.
We run each configuration 10 times, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect enough workgroups to launch the non-cooperative
kernel, and the non-cooperative kernel execution time.
%
%We sanity-check
%whether all kernels perform their computations correctly by comparing
%their result with reference results.

@@

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (median over all
applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (median over all applications
and inputs). We use the median as our data included some drastic
outliers; these outliers appear to happen with applications/inputs in
which the resizing barrier are not called very frequently and the
graphics task asks for half or more of the resources. For example, a
medium graphics workload for bfs on the rmat input has over an
8$\times$ overhead when asking for all but one of the workgroups for
graphics. As \myfiglong\ref{fig:performance} shows, most of our
applications/inputs are much better behaved than this. Future work
includes examining these problem applications/inputs in more detail,
possibly to insert more calls to $\offerfork$ and $\offerkill$.

@@

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 ms. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a median period of 104 ms). If half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload achieves
its goal period (median of 40 ms). 
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time of the cooperative kernel.

Still, heavy workloads meet their period by allocating half
non-cooperative workgroups, incurring a slow down of less than
1.5$\times$ (median). Light and medium workloads meet their period
with only a small overhead of the cooperative kernel; 
1.04$\times$ and 1.08$\times$ median slowdown respectively.

@@

\subsection{Comparison with Task Level Preemption}\label{sec:nvidiacomparison}

@@

Some recent GPUs, including Nvidia's Pascal architecture, provide
hardware support for instruction-level preemption.  Few details of how
this works are available, but from reading Nvidia documentation
and online material~\cite{PascalWhitepaper,anandtech}, and running
litmus tests on a Pascal platform, our understanding is that
preemption of entire kernels, but not of individual workgroups running
a kernel is supported.  Intel GPUs do
not provide this feature, and out OpenCL 2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making a direct comparison impossible.
We present here a theoretical analysis of the overheads associated with
sharing the GPU between graphics and compute tasks via kernel-level
preemption.

@@

We have proposed \emph{cooperative kernels}, a small set of GPU
programming extensions that allow long-running, blocking kernels to be
fairly scheduled and to share GPU resources with other workloads.
Experimental results using our megakernel-based prototype \cutthree{over three Intel GPUs}
shows that the model is a good fit for the needs of current
GPU-accelerated irregular algorithms.  The
performance that could be gained through a native implementation with
dedicated driver support would be even better.
%
Avenues for future work include seeking
additional classes of irregular algorithms to which the model might
(be extended to) apply (to), investigating implementing native support
in open source drivers, and integrating
cooperative kernels into template- and compiler-based programming
models for graph algorithms on
GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.
%  As more vendors
%adopt the advanced features of OpenCL that our approach requires, we
%will be in a position to assess the programming model across a wider
%range of devices.

>>>>

We discuss our experience porting irregular algorithms to cooperative
kernels and describe the GPUs on which we evaluate these applications
(\mysec\ref{sec:portingalgorithms}).  For these GPUs, we report on
experiments to determine non-cooperative workloads that model the
requirements of various graphics rendering tasks
(\mysec\ref{sec:sizingnoncoop}).  We then examine the overhead
associated with moving to cooperative kernels when multitasking is
\emph{not} required (\mysec\ref{sec:overhead}), as well as the
responsiveness and throughput observed when a cooperative kernel is
multi-tasked with non-cooperative workloads
(\mysec\ref{sec:responsiveness}). Finally, we compare
against a performance model of \emph{kernel-level} preemption, which
we understand to be what current Nvidia GPUs provide (\mysec\ref{sec:nvidiacomparison}).

%To our knowledge, there are no competing methods against which to
%meaningfully compare our approach.  We are not aware of any existing
%platform that supports multitasking between long-running, fully-occupant
%OpenCL kernels.  Very recent \nvidia{} GPUs, to which we do not yet have
%access, support preemption for graphics
%applications~\cite{PascalWhitepaper}, but our understanding is that this
%cannot yet be leveraged from CUDA, and certainly not from OpenCL.

@@

We devised the following procedure to determine non-cooperative OpenCL
workloads that simulate the computational intensity of various
graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgroups of a GPU
for a parameterised time period $t$, invoked in an infinite loop by a
host application.  We then searched for a maximum value for $t$ that
allowed the synthetic kernel to execute without having an observable
impact on graphics rendering.  Using the computed value, we ran the
host application for $X$ seconds, measuring the time $Y < X$ dedicated
to GPU execution during this period and the number of kernel launches
$n$ that were issued.  We used $X \geq 10$ in all experiments.  The
values $(X-Y)/n$ and $X/n$ estimate the average time spent using the
GPU to render the display between kernel calls (call this $E$) and the
period at which the OS requires the GPU for display rendering (call
this $P$), respectively.

@@

Our results are the following: $P=70ms$ and $E=3ms$ for light;
$P=40ms$, $E=3ms$ for medium; and $P=40ms$, $E=10ms$ for heavy. For
medium and heavy, the 40ms period coincides with the human persistence
of vision. The 3ms execution duration of both light and medium
configurations indicates that GPU computation is cheaper for basic
display rendering compared with more complex rendering that uses
shaders. These results guide our evaluation of cooperative scheduling
(\mysec\ref{sec:responsiveness}).

@@

Our experiments span 11 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of
workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 240 configurations.
We run each configuration 10 times, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect workgroups to launch the non-cooperative kernel,
and the non-cooperative kernel execution time.
%
%We sanity-check
%whether all kernels perform their computations correctly by comparing
%their result with reference results.

@@

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (median over all
applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (median over all applications
and inputs). We use the median as our data included some outliers;
these outliers happen with applications/inputs in which the resizing
barrier are not called very frequently and the graphics task requires
half or more workgroups. For example, a medium graphics workload for
bfs on the rmat input has over an 8$\times$ overhead when asking for
all but one of the workgroups. As \myfiglong\ref{fig:performance}
shows, most of our applications/inputs are much better behaved than
this. Future work will examine these problem applications/inputs in
more detail, possibly inserting more calls to $\offerfork$ and
$\offerkill$.

@@

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 ms. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a median period of 104 ms). If half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload achieves
its goal period (median of 40 ms). 
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time of the cooperative kernel.
%
Still, heavy workloads meet their period by allocating half
non-cooperative workgroups, incurring a slow down of less than
1.5$\times$ (median). Light and medium workloads meet their period
with only a small overhead of the cooperative kernel; 
1.04$\times$ and 1.08$\times$ median slowdown respectively.

@@

\subsection{Comparison with Kernel-Level Preemption}\label{sec:nvidiacomparison}

@@

Some recent GPUs, including Nvidia's Pascal architecture, provide
hardware support for instruction-level preemption.  Few details of how
this works are available, but from reading Nvidia documentation and
online material~\cite{PascalWhitepaper,anandtech}, our understanding
is that preemption of entire kernels, but not of individual workgroups
is supported.  Intel GPUs do not provide this feature, and out OpenCL
2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making
a direct comparison impossible.  We present here a theoretical
analysis of the overheads associated with sharing the GPU between
graphics and compute tasks via kernel-level preemption.

@@

We have proposed \emph{cooperative kernels}, a small set of GPU
programming extensions that allow long-running, blocking kernels to be
fairly scheduled and to share GPU resources with other workloads.
Experimental results using our megakernel-based prototype shows that
the model is a good fit for current GPU-accelerated irregular
algorithms.  The performance that could be gained through a native
implementation with driver support would be even better.
%
Avenues for future work include seeking additional classes of
irregular algorithms to which the model might (be extended to) apply
(to), investigating implementing native support in open source
drivers, and integrating cooperative kernels into template- and
compiler-based programming models for graph algorithms on
GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.
%  As more vendors
%adopt the advanced features of OpenCL that our approach requires, we
%will be in a position to assess the programming model across a wider
%range of devices.
